Repeated loss of variation in insect ovary morphology highlights the role of developmental constraint in life-history evolution

### 2021-01-18 manuscript.Rmd
adding conda environment; rerunning analyses; updating manuscript and README

The number of offspring an organism can produce is a key component of its evolutionary fitness and life-history. Here we perform a test of the hypothesized trade off between the number and size of offspring using thousands of descriptions of the number of egg-producing compartments in the insect ovary (ovarioles), a common proxy for potential offspring number in insects. In contrast to prior claims, we find that ovariole number is not generally negatively correlated with the size of insect eggs, and we highlight several factors that may have contributed to this size-number trade off being strongly asserted in previous studies. We reconstruct the evolutionary history of the nurse cell arrangement within the ovariole, and show that the diversification of ovariole number and egg size have both been largely independent of nurse cell presence or position within the ovariole. Instead we show that ovariole number evolution has been shaped by a series of transitions between variable and invariant states, with multiple independent lineages evolving to have almost no variation in ovariole number. We highlight the implications of these invariant lineages on our understanding of the specification of ovariole number during development, as well as the importance of considering developmental processes in theories of life-history evolution.

@@

We combined the data we collected on total ovariole number with existing datasets of egg size and shape[@church2019dataset], insect lifetime fecundity and dry adult body weight[@Gilbert2007], average adult body length per insect family[@rainford2016phylogenetic], several lineage-specific measures of adult body size[@iwata1966large;@Iwata1966a;@waloff1954number;@starmer2003phylogenetic;@reinhardt2005low], and the mode of oogensis[@buning1994insect]. See supplementary methods section \ref{sec:combining} for details.

@@

Ancestral state reconstruction of oogenesis mode was performed with the R package corHMM[@beaulieu2013identifying], version `r packageVersion('corHMM')`, and models of trait evolution were compared using the R package Ouwie[@beaulieu2012modeling], version `r packageVersion('OUwie')`. Ancestral state reconstruction and model comparison were repeated 100 times over a posterior distribution of trees and resampling data to account for variation across records for the same taxon. See supplementary methods section \ref{sec:oogenesis-model}.

Other comparisons of model fit were performed using the R package geiger[@harmon2007geiger], version `r packageVersion('geiger')`, and validated using a parametric bootstrap with the R package arbutus[@pennell2015model], version `r packageVersion('arbutus')`. See supplementary methods section \ref{sec:bootstrap}.

Analyses of evolutionary rate were performed using BAMM[@rabosky2014automatic], version 2.5.0. For this analysis, we calculated the average ovariole number (log~10~ transformed) for each genus present in the phylogeny (507 taxa). We used the R package BAMMtools (version `r packageVersion('BAMMtools')`)[@rabosky2014bamm] to select appropriate priors, and ran BAMM for the maximum number of generations ($2*10^{-9}$), sampling every 10^6^ generations. Convergence was evaluated both visually (Fig. \ref{fig:BAMM-analysis-convergence}) and numerically. Running BAMM for the maximum possible number of generations and selecting the optimum burn-in\ref{fig:BAMM-analysis-burnin} resulted in an effective size for the number of shifts of `r round(es_nshifts,2)`, and for log-likelihood of `r round(es_logl,2)`. Repeated BAMM analyses showed similar distributions of high and low rate regimes, indicating the implications for ovariole number evolution are robust to uncertainty in rate estimates. See supplementary methods section \ref{sec:BAMM} for details.

@@

The dataset of insect ovariole number is available at Dryad (doi:10.5061/dryad.59zw3r253). The code and phylogenetic trees required to reproduce all the analyses, figures, and generate the manuscript files are provided at 'https://github.com/shchurch/insect_ovariole_number_evolution_2020', commit 6cf446a.

@@

All regressions were performed over a posterior distribution of trees and accounting for phenotypic uncertainty; therefore these analyses produced a range of p-values. We considered a relationship significant when the maximum p-value observed from this range was below the threshold 0.01. We report the maximum p-value throughout this manuscript, and have included the full range in the supplementary information (see Table \ref{tab:PGLS-posterior}).

@@

Ovariole number has been hypothesized to be negatively correlated with egg size[@montague1981reproductive;@berrigan1991allometry;@starmer2003phylogenetic]. This hypothesis is based on the predictions that (1) female reproduction is resource-limited, therefore egg size should trade off with egg number, and (2) ovariole number can serve as a proxy for egg number[@montague1981reproductive;@smith1974optimal]. Our data do not support the hypothesis of a negative relationship between egg size and ovariole number across insects (p-value `r all_table[2,]$'pval_max'`; Figs. 2a, \ref{fig:PGLS-egg-ON-genus}, Table \ref{tab:PGLS-posterior}).

Given that this prediction is often conditioned on taking into account adult body size[@stewart1991reproductive;@starmer2003phylogenetic], we combined data on ovariole number and egg size with several measurements of body size, and applied a size correction that accounts for phylogenetic relationships [@revell2009size]. Across insects we did not observe a significant relationship between ovariole number and egg size when controlling for body size (p-value `r all_table[6,]$'pval_max'`; Fig. 2b, \ref{fig:PGLS-Gilbert-name}, \ref{fig:PGLS-egg-ON-family}, Table \ref{tab:PGLS-posterior}), though we do note the presence of a non-significant negative trend, primarily driven by a few extreme values (e.g. wasps in the family Trigonalidae, which have many ovarioles and lay minute eggs, Fig. \ref{fig:PGLS-egg-ON-family}).

We assessed the relationship between these traits for four sublineages of insects and found that across Drosophilidae species, egg size is indeed strongly negatively correlated with ovariole number when accounting for body size (p-value <0.001; Fig. 2c, Table \ref{tab:PGLS-posterior}). However this relationship did not hold for any of the three other lineages we analyzed (e.g. Orthoptera p-value `r all_table[9,]$'pval_max'`; Fig. 2d, \ref{fig:PGLS-Coleoptera-Hymenoptera}, Table \ref{tab:PGLS-posterior}).

We additionally tested whether ovariole number is positively correlated with adult body size, and in contrast to previous studies[@honvek1993intraspecific], we found no correlation between ovariole number and adult body weight or length across insects (p-values `r body_all_table[3,]$'pval_max'` and `r body_all_table[1,]$'pval_max'`, respectively; Fig. \ref{fig:PGLS-body-plot}, Table \ref{tab:PGLS-body-posterior}).

![**Tests of the hypothesized trade off between egg size and ovariole number.** a, Egg volume (mm^3^) and ovariole number, both log~10~ scale, n=`r nrow(egg_ON_name_pgls)` species. See section _Modeling ovariole number evolution_ for discussion of the enrichment of certain low values of ovariole number (points appearing vertically arranged) b, Egg volume and ovariole number, residuals to adult dry body weight, n=`r length(which(egg_ON_gilbert_genus_pgls$rank %in% tree$tip.label))` genera. c, Drosophilidae egg volume and ovariole number, residuals to thorax length, n=`r length(which(DIP_pgls$rank %in% dros_tree$tip.label))` species. d, Orthoptera egg volume and ovariole number, residuals to body length, n=`r length(which(ORT_pgls$rank %in% tree$tip.label))` genera.](manuscript_figures/Fig_Egg_V2-01.png)

@@

Using a previously reported dataset of lifetime fecundity measurements across insects[@Gilbert2007], we assessed the relationship between lifetime fecundity and ovariole number, and observed a positive, but not statistically significant, correlation (p-value `r all_table[8,]$'pval_max'`; Figs. 3, \ref{fig:PGLS-fecundity-name}). We interpret our results, in conjunction with those previously reported, to suggest that ovariole number, when variable across insects in a lineage, may be one factor among many influencing the number of eggs produced. However, we caution against using ovariole number as a direct mathematical proxy for offspring number.

![**The relationship between lifetime fecundity and ovariole number.** Both values are shown on a log~10~ scale. n=`r length(na.omit(fecundity_genus_pgls$fecundity))` genera, colored according to the groups shown in Fig. 1b.](manuscript_figures/Fig_Fecundity_V1-01.png)

@@

In addition to the number of ovarioles, insect ovary morphology has been classified into several modes of oogenesis based on the presence and position of nutritive cells called nurse cells[@buning1994insect], (Fig. 4a). The ovary of the well-studied species _D. melanogaster_ is an example of a meroistic oogenesis mode, meaning that its ovarioles contain nurse cells of germ line origin that are connected to developing oocytes via cytoplasmic bridges[@KingRobertC1970OdiD]. In insects with a polytrophic arrangement, these nurse cells are clonally related and immediately adjacent to each oocyte. An alternative arrangement is seen in telotrophic meroistic ovaries, where oocytes in each ovariole are connected to a common pool of nurse cells located in the germarium[@buning1994insect]. Meroistic ovaries are thought to have evolved from a panoistic ancestral state, meaning they lack nurse cells[@buning1994insect]. Using a previously published set of descriptions of these oogenesis modes across insects[@buning1994insect], we reconstructed the evolutionary transitions between these states. Consistent with previous analyses[@buning1994insect], we found that the ancestral insect likely had panoistic ovaries (without nurse cells), with several independent shifts to both telotrophic and polytrophic meroistic modes, and at least two reversals from meroistic back to panoistic (Figs. 4b, \ref{fig:reconstruct-ovary}).

@@

Using the dataset compiled here and a previously published phylogeny of insects (Fig. 5a)[@church2019insect], we modeled the rate of evolutionary change in ovariole number (Figs. \ref{fig:arbutus}, \ref{fig:BAMM-analysis-convergence}, \ref{fig:BAMM-analysis-burnin}, \ref{fig:BAMM-analysis-results}). We observed substantial rate heterogeneity in the evolution of ovariole number (Fig. \ref{fig:BAMM-analysis-results}), meaning that for some lineages ovariole number has evolved rapidly where in others, ovariole number has evolved very slowly or not at all. The most striking example of this are the multiple lineages which have independently evolved invariant or near-invariant ovariole number across taxa (e.g. nearly all Lepidoptera have exactly eight ovarioles, Fig. 5b, cyan), from an ancestral variable state. These invariant lineages were objectively identified by finding regions of the phylogeny that experience extremely low rates of ovariole number diversification (Figs. \ref{fig:BAMM-analysis-results}, \ref{fig:BAMM-anlaysis-mean-rates}). Using this approach, we found that invariant ovariole numbers have evolved at least nine times independently across insects, with several subsequent reversals from invariant to variable states (Fig. 5a).

We find that the rate of evolutionary change in ovariole number is correlated with the number of ovarioles: lineages with relatively low ovariole number also experience relatively low degrees of change (Fig. \ref{fig:arbutus}). This is evidenced by the fact that, of the nine invariant lineages, none have greater than seven ovarioles per ovary (Fig. 5c). However we note that not all insects with low ovariole counts are in invariant lineages; many insects with fewer than 14 total ovarioles are in lineages with relatively high rates of intra- and interspecific ovariole number variation (Fig. 5)

@@

When considering the developmental processes that could give rise to invariant ovariole number, we propose that the major determinants of ovariole number known from _Drosophila_ may not apply. To achieve an invariant ovariole number, these processes might instead include mechanisms for strict counting of individual cells or discrete cell subpopulations. In the former, if the cells that ultimately comprised a terminal filament were derived by mitotic division from a single progenitor, rather than by cellular rearrangements as is the case in _Drosophila_[@godt1995mechanisms], then an invariant ovariole number could be achieved via strict control of the number of precursor cells. Alternatively, an invariant ovariole number could be achieved by partitioning the starting population of precursor cells into a tightly regulated number of subpopulations. This would again be a departure from known mechanisms in _Drosophila_, in which a variable number of precursor cells are gathered into terminal filaments until the population is depleted[@godt1995mechanisms;@sarikaya2012roles]. The determining factor for partitioning the precursor pool could be, for example, a spatially variable morphogen emanating from adjacent tissues [@lawrence1996morphogens] or a reaction-diffusion patterning process [@kondo2010reaction] within the developing ovary, as has been shown to generate fixed numbers of multicellular structures in other developmental contexts[@salazar2008tooth;@hatini2001divide;@clark2019arthropod]. These predictions could be tested by characterizing the dynamics of cell number and position across invariant lineages, and making comparisons to corresponding data from their variable relatives.

>>>>

The number of offspring an organism can produce is a key component of its evolutionary fitness and life-history. Here we perform a test of the hypothesized trade off between the number and size of offspring using thousands of descriptions of the number of egg-producing compartments in the insect ovary (ovarioles), a common proxy for potential offspring number in insects. We find evidence of a negative relationship between egg size and ovariole number when accounting for adult body size. However in contrast to prior claims, we note that this relationship is not generalizable across all insect clades, and we highlight several factors that may have contributed to this size-number trade off being stated as a general rule in previous studies. We reconstruct the evolutionary history of the nurse cell arrangement within the ovariole, and show that the diversification of ovariole number and egg size have both been largely independent of nurse cell presence or position within the ovariole. Instead we show that ovariole number evolution has been shaped by a series of transitions between variable and invariant states, with multiple independent lineages evolving to have almost no variation in ovariole number. We highlight the implications of these invariant lineages on our understanding of the specification of ovariole number during development, as well as the importance of considering developmental processes in theories of life-history evolution.

@@

We combined the data we collected on total ovariole number with existing datasets of egg size and shape[@church2019dataset], insect lifetime fecundity and dry adult body mass[@Gilbert2007], average adult body length per insect family[@rainford2016phylogenetic], several lineage-specific measures of adult body size[@iwata1966large;@Iwata1966a;@waloff1954number;@starmer2003phylogenetic;@reinhardt2005low], and the mode of oogenesis[@buning1994insect]. See supplementary methods section \ref{sec:combining} for details.

@@

For two regressions comparing egg size to ovariole number while accounting for adult body size, we tested alternative hypotheses of evolution by simulating new data. We considered two such hypotheses: no evolutionary correlation with ovariole number, and a strong correlation with ovariole number (slope of -1). For each trait we simulated 1,000 datasets using evolutionary parameters fit under a Brownian Motion model in the R packages geiger (version `r packageVersion('geiger')`)[@revell2010phylogenetic], and phylolm (version `r packageVersion('phylolm')`)[@tung2014linear].

Ancestral state reconstruction of oogenesis mode was performed with the R package corHMM (version 1.22)[@beaulieu2013identifying], and models of trait evolution were compared using the R package Ouwie (version 1.57)[@beaulieu2012modeling]. Ancestral state reconstruction and model comparison were repeated 100 times over a posterior distribution of trees and resampling data to account for variation across records for the same taxon. See supplementary methods section \ref{sec:oogenesis-model}.

Other comparisons of model fit were performed using the R package geiger(version `r packageVersion('geiger')`)[@harmon2007geiger] and validated using a parametric bootstrap with the R package arbutus (version `r packageVersion('arbutus')`)[@pennell2015model]. See supplementary methods section \ref{sec:bootstrap}.

Analyses of evolutionary rate were performed using BAMM (version 2.5.0)[@rabosky2014automatic]. For this analysis, we calculated the average ovariole number (log~10~ transformed) for each genus present in the phylogeny (507 taxa). We used the R package BAMMtools (version `r packageVersion('BAMMtools')`)[@rabosky2014bamm] to select appropriate priors, and ran BAMM for the maximum number of generations ($2*10^{-9}$), sampling every 10^6^ generations. Convergence was evaluated both visually (Fig. \ref{fig:BAMM-analysis-convergence}) and numerically. Running BAMM for the maximum possible number of generations and selecting the optimum burn-in (Fig. \ref{fig:BAMM-analysis-burnin}) resulted in an effective size for the number of shifts of `r round(es_nshifts,2)`, and for log-likelihood of `r round(es_logl,2)`. Repeated BAMM analyses showed similar distributions of high and low rate regimes, indicating the implications for ovariole number evolution are robust to uncertainty in rate estimates. See supplementary methods section \ref{sec:BAMM} for details.

@@

The dataset of insect ovariole number is available at Dryad, doi:10.5061/dryad.59zw3r253.

## Code availability {-}

The code and phylogenetic trees required to reproduce all the analyses, figures, and generate the manuscript files are provided at 'https://github.com/shchurch/insect_ovariole_number_evolution_2020', commit 6cf446a. All analyses performed in R (version 4.0.3) were done so in a clean environment, built with conda (version 4.9.2), and instructions for rebuilding this environment are provided in the same repository.

@@

All phylogenetic regressions were performed using both the maximum clade credibility (MCC) tree. We considered a relationship significant when the MCC p-value was below the threshold 0.01. To assess the robustness of results to uncertainty in phylogenetic relationships, we also report the number of regressions over the posterior distribution of phylogenetic trees that gave a significant result (see Table \ref{tab:PGLS-posterior-egg-ON}).

For two comparisons, we validated that our tests had sufficient statistical power using the selected threshold by comparing the distribution of p-values from regressions of observed data to regressions of data simulated under alternative hypotheses. We compared the results of analyses of our observed to those based on simulated data to evaluate the likelihood of false positives (comparing to data simulatd under no correlation) and false negatives (comparing to data simulated with strong correlation).

@@

Ovariole number has been hypothesized to be negatively correlated with egg size[@montague1981reproductive;@berrigan1991allometry;@starmer2003phylogenetic]. This hypothesis is based on the predictions that (1) female reproduction is resource-limited, therefore egg size should trade off with egg number, and (2) ovariole number can serve as a proxy for egg number[@montague1981reproductive;@smith1974optimal]. We did not observe a significant negative relationship when comparing egg size and ovariole number across insect species (Fig. 2a, Table \ref{tab:PGLS-posterior-egg-ON}, p-value `r round(pgls_results_egg_ON_name$tTable[[8]],3)`, n=`r pgls_results_egg_ON_name$dims$N`). We also compared egg size and ovariole number across insect genera and did not observe a significant relationship (Fig. \ref{fig:PGLS-egg-ON-genus}, p-value `r round(pgls_results_egg_ON_genus$tTable[[8]],3)`, n=`r pgls_results_egg_ON_genus$dims$N`).

Given that this predicted relationship is often conditioned on taking into account adult body size[@stewart1991reproductive;@starmer2003phylogenetic], we combined data on ovariole number and egg size with several measurements of body size, and applied a size correction that accounts for phylogenetic relationships[@revell2009size]. When accounting for adult body mass, we observed a significant negative relationship between egg size and ovariole number (Fig. 2b, \ref{fig:PGLS-Gilbert-name}, p-value `r round(pgls_results_gilbert_genus$tTable[[8]],3)`, n=`r pgls_results_gilbert_genus$dims$N`). To evaluate the robustness of this result, we repeated the analysis 1000 times, taking into account uncertainty in both the phylogeny and trait measurements. Out of 1000 regressions, `r sum(sapply(allometry_gilbert_genus,function(x){x$tTable[[8]]}) < 0.01)` indicated a significant negative relationship (Table \ref{tab:PGLS-posterior-egg-ON}). We performed the same comparison accounting for adult body length, and likewise observed a significant negative relationship (Fig. \ref{fig:PGLS-family-length}, p-value `r gsub(0,"<0.001",round(pgls_results_body_egg_ON_family$tTable[[8]],3))`, n=`r pgls_results_body_egg_ON_family$dims$N`), supported by `r sum(sapply(allometry_body_egg_ON_family,function(x){x$tTable[[8]]}) < 0.01)` of 1000 repeated analyses (Table \ref{tab:PGLS-posterior-egg-ON}).

We further explored these results using two methods: First, to evaluate our findings against alternative evolutionary hypotheses, we compared these results to regressions based on simulated data. Our results showed that when considering body size, the slope of the regression of egg size and ovariole number is lower than we would expect to observe by chance, as assessed by comparing to data simulated with no evolutionary correlation (Fig. \ref{fig:simulated}). However, for both adult body length and dry mass, the slope of the regressions on observed data are not within the range that would be expected under an strong negative correlation (slope of -1 in log-log space, Fig. \ref{fig:simulated}). This suggests the presence of a weak evolutionary relationship bewteeen ovariole number and egg size, when accounting for body size.

Second, we assessed the relationship between egg size and ovariole number, accounting for body size, within four subclades of insects. We found that across Drosophilidae fly species, egg size is indeed strongly negatively correlated with ovariole number when accounting for body size (Fig. 2c, Table \ref{tab:PGLS-posterior-clade}, p-value `r gsub(0,"<0.001",round(pgls_results_DIP$tTable[[8]],3))`, n=`r pgls_results_DIP$dims$N`). For grasshoppers and crickets (Orthoptera), beetles (Coleoptera), and wasps (Hymenoptera), we observed no significant relationship between ovariole number and egg size, even when accounting for body size (Fig. 2d, Table \ref{tab:PGLS-posterior-clade}, \ref{fig:PGLS-Coleoptera-Hymenoptera}, Orthoptera: p-value `r round(pgls_results_ORT$tTable[[8]],3)`, n=`r pgls_results_ORT$dims$N`, Coleoptera: p-value `r round(pgls_results_COL1$tTable[[8]],3)`, n=`r pgls_results_COL1$dims$N`, Hymenoptera: p-value `r round(pgls_results_HYM$tTable[[8]],3)`, n=`r round(pgls_results_HYM$dims$N)`). This indicates that, while a strong negative correlation between egg size and ovariole number exists for some insects, it does not represent a universal pattern across insect clades.

Finally, we tested whether ovariole number is positively correlated with adult body size, and in contrast to previous studies[@honvek1993intraspecific], we found no correlation between ovariole number and adult body mass or length across insects (Fig. \ref{fig:PGLS-body-plot}, Table \ref{tab:PGLS-posterior-body}, body mass: p-value `r round(pgls_results_body_gilbert_name$tTable[[8]],3)`, n=`r pgls_results_body_gilbert_genus$dims$N`, body length: p-value `r round(pgls_results_body_ON_family$tTable[[8]],3)`, n=`r pgls_results_body_ON_family$dims$N`). Of the four subclades considered, only insects in the order Orthoptera had a positive relationship between body size and ovariole number (Table \ref{tab:PGLS-posterior-body}, p-value `r round(pgls_results_body_ORT$tTable[[8]],3)`, n=`r pgls_results_body_ORT$dims$N`).

![**Tests of the hypothesized trade off between egg size and ovariole number.** a, Egg volume (mm^3^) and ovariole number, both log~10~ scale; points represent insect species. See section _Modeling ovariole number evolution_ for discussion of the enrichment of certain low values of ovariole number (points appearing vertically arranged) b, Egg volume and ovariole number, residuals to dry adult body mass, points represent genera. c, Drosophilidae egg volume and ovariole number, residuals to thorax length, points represent species. d, Orthoptera egg volume and ovariole number, residuals to body length, points represent genera.](manuscript_figures/Fig_Egg_V2-01.png)

@@

Using a previously reported dataset of lifetime fecundity measurements across insects[@Gilbert2007], we assessed the relationship between lifetime fecundity and ovariole number. We observed a significant positive relationship (Fig. 3, p-value `r round(pgls_results_fecundity_genus$tTable[[8]],3)`, n=`r pgls_results_fecundity_genus$dims$N`), however, a substantial fraction of repeat analyses show these results are not robust to uncertainty (`r sum(sapply(allometry_fecundity_genus,function(x){x$tTable[[8]]}) > 0.01)` of 1000 regressions are not significant, Table \ref{tab:PGLS-posterior-fecundity}). We note that this relationship is largely defined by the absence of insects with high ovariole number and low fecundity (Fig. 3, empty bottom right corner), while for insects with low ovariole number, fecundity varied over more than three orders of magnitude. We interpret our results, in conjunction with those previously reported, to suggest that ovariole number, when variable across insects in a lineage, may be one factor among many influencing the number of eggs produced. However, we caution against using ovariole number as a direct mathematical proxy for offspring number.

![**The relationship between lifetime fecundity and ovariole number.** Both values are shown on a log~10~ scale. Points represent insect genera and are colored according to the groups shown in Fig. 1b.](manuscript_figures/Fig_Fecundity_V1-01.png)

@@

In addition to the number of ovarioles, insect ovary morphology has been classified into several modes of oogenesis based on the presence and position of nutritive cells called nurse cells[@buning1994insect], (Fig. 4a). Egg formation in the well-studied species _D. melanogaster_ is an example of a meroistic oogenesis mode, meaning that its ovarioles contain nurse cells of germ line origin that are connected to developing oocytes via cytoplasmic bridges[@KingRobertC1970OdiD]. In insects with a polytrophic arrangement, these nurse cells are clonally related and immediately adjacent to each oocyte. An alternative arrangement is seen in telotrophic meroistic ovaries, where oocytes in each ovariole are connected to a common pool of nurse cells located in the germarium[@buning1994insect]. Meroistic ovaries are thought to have evolved from an ancestral panoistic mode, meaning they lack nurse cells[@buning1994insect]. Using a previously published set of descriptions of these oogenesis modes across insects[@buning1994insect], we reconstructed the evolutionary transitions between these states. Consistent with previous analyses[@buning1994insect], we found that the ancestral insect likely had panoistic ovaries (lacking nurse cells), with several independent shifts to both telotrophic and polytrophic meroistic modes, and at least two reversals from meroistic back to panoistic (Figs. 4b, \ref{fig:reconstruct-ovary}).

@@

Using the dataset compiled here and a previously published phylogeny of insects (Fig. 5a)[@church2019insect], we modeled the rate of evolutionary change in ovariole number (Figs. \ref{fig:arbutus}, \ref{fig:BAMM-analysis-convergence}, \ref{fig:BAMM-analysis-burnin}, \ref{fig:BAMM-analysis-results}). We observed substantial rate heterogeneity in the evolution of ovariole number (Fig. \ref{fig:BAMM-analysis-results}), meaning that for some lineages ovariole number has evolved rapidly where in others, ovariole number has evolved very slowly or not at all. The most striking example of this are the multiple lineages which have independently evolved invariant or near-invariant ovariole number across taxa (e.g. nearly all Lepidoptera have exactly eight ovarioles, Fig. 5b, cyan), from an ancestral variable state. These invariant lineages were identified by finding regions of the phylogeny that experience extremely low rates of ovariole number diversification (Figs. \ref{fig:BAMM-analysis-results}, \ref{fig:BAMM-anlaysis-mean-rates}). Using this approach, we found that invariant ovariole numbers have evolved at least nine times independently across insects, with several subsequent reversals from invariant to variable states (Fig. 5a).

We find that the rate of evolutionary change in ovariole number is correlated with the number of ovarioles: lineages with relatively low ovariole number also experience relatively low degrees of ovariole number change (Fig. \ref{fig:arbutus}). This is evidenced by the fact that, of the nine invariant lineages, none have greater than seven ovarioles per ovary (Fig. 5c). However we note that not all insects with low ovariole counts are in invariant lineages; many insects with fewer than 14 total ovarioles are in lineages with relatively high rates of intra- and interspecific ovariole number variation (Fig. 5)

@@

When considering the developmental processes that could give rise to invariant ovariole number, we propose that the major determinants of ovariole number known from _Drosophila_ may not apply. To achieve an invariant ovariole number, these processes might instead include mechanisms for strict counting of individual cells or discrete cell subpopulations. In the former, if the cells that ultimately comprised a terminal filament were derived by mitotic division from a single progenitor, rather than by cellular rearrangements as is the case in _Drosophila_[@godt1995mechanisms], then an invariant ovariole number could be achieved via strict control of the number of precursor cells. Alternatively, an invariant ovariole number could be achieved by partitioning the starting population of precursor cells into a tightly regulated number of subpopulations. This would again be a departure from known mechanisms in _Drosophila_, in which a variable number of precursor cells are gathered into terminal filaments until the population is depleted[@godt1995mechanisms;@sarikaya2012roles]. The determining factor for partitioning the precursor pool could be, for example, a spatially variable morphogen emanating from adjacent tissues [@lawrence1996morphogens] or a reaction-diffusion patterning process [@kondo2010reaction] within the developing ovary, as these have been shown to generate fixed numbers of multicellular structures in other developmental contexts[@salazar2008tooth;@hatini2001divide;@clark2019arthropod]. These predictions could be tested by characterizing the dynamics of cell number and position across invariant lineages, and making comparisons to corresponding data from their variable relatives.
